Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-10-02 content.tex
Fixed the circularity issue with unroll after talking with Mitch.

%%
There is a systematic process for transforming off-the-shelf programming language semantics into a form amenable to \emph{regular} analysis that has been widely applied with great success to production programming languages ~\citet{dvanhorn:VanHorn2010Abstracting} (a technique called abstracting abstract machines, or AAM).
%
A contribution of this paper is a systematic process to construct \emph{pushdown} analyses of programming languages, due to the precision (and often performance) benefits.
%%

@@

The remaining sections of the paper are
\begin{itemize}
\item{section \ref{sec:pdcfa}: we derive a cousin of PDCFA}
\item{section \ref{sec:cfa2}: we make additions to the previous semantics to get a direct-style CFA2 without first-class control}
\item{section \ref{sec:sr}: we give a novel analysis of delimited and composable first-class control.}
\end{itemize}

@@

%%
The tail of a continuation in the context of a function will contain the stack-less context in which the function was called, in order to link up with the proper call-site(s).
%
The context includes the function (or unique label of the function), the environment, and the store.
%
Notice that this choice can be viewed as a special allocation strategy for continuations in the AAM viewpoint, but we separate the table of continuations from the store since continuations themselves contain stores in $\rt$ frames --- this leads to a recursive data-structure that we are trying to avoid.
%
The store is an important ingredient to maintaining enough precision to keep a pushdown abstraction.
%
The semantic spaces for the machine are modified thusly:

@@

The $\mathit{unroll}$ function interprets what are all the valid continuations that $\mktab$ encodes for a given continuation that contains $\krt{\mctx}$:
\begin{mathpar}
  \inferrule{ }{\kmt \in \unroll{\mktab}{\kmt}} \quad
  \inferrule{\mkont \in \unroll{\mktab}{\mkont'}}{\mkframe:\mkont \in \unroll{\mktab}{\mkframe:\mkont'}} \quad
  \inferrule{\mkont^\circ \in \mktab(\mctx) \\
             \mkont \in \unroll{\mktab}{\mkont^\circ}}
            {\mkont \in \unroll{\mktab}{\krt{\mctx}}}
\end{mathpar}

The astute reader may find this definition is dubious, since there is a possibility for unguarded corecursion here.
%
A function with ill-founded recursion can lead to a table such that $\krt{\mctx} \in \mktab(\mctx)$.
%
The abstractions that we impose on our data might also cause this situation, even though the concrete execution might always terminate.
%
Our interpretation of $\stepto$ is in a fixed-point computation, so circularities are not a problem for termination.

@@

\section{Deriving GC from the introspective pushdown model}

@@

@@

CFA2 is the first published analysis of a higher-order programming language that could properly match calls and returns.
%
We will show that it fits well into the same presentation we gave for PDCFA.
%
Vardoulakis and Shivers had a clear goal of harnessing the extra information a pushdown model provides to produce a high-precision analysis that works well in practice.
%
This resulted in more than just the call/return matching of the previous section, which is why we are showing the two separately.
%
There are two orthogonal features of the semantics:
\begin{enumerate}
\item{stack allocation for some bindings in an additional $\mframe \in \Store$}
\item{strong updates on stack frames for resolved non-determinism}
\end{enumerate}
The first of these is an addition to the stack-less context.
%
There is a conservative pre-analysis that checks locally whether a binding will never escape, and classifies references (labeled with a distinguishing $\mlab$ from an arbitrary space of labels) as able to use the stack frame or not.
%
Their criteria for a binding never escaping is that it is never referenced in a function that is not its binder.
%
This can be extended in a language with more linguistic features; see Kranz's thesis about register-allocatable bindings in the Orbit Scheme compiler \citep{ianjohnson:kranz:thesis:1988}.
%
A stackable reference is one that appears in the body of the binding function, by which we mean not within the body of a nested function.
%
They use the information that a binding never escapes to not bother allocating it in the heap.
%
This has the advantage of not changing the heap, and thus leads to less propagation.
%
The addition of these stack frames makes the analysis exponential in theory, though in practice they have been observed to decrease running time in most cases.
%%

@@

%%
We show only the significantly modified rules of the semantics in figure \ref{fig:frame-semantics}.
%
The other rules simply carry along the extra $\mframe$ component untouched.
\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar[\mlab]{\mvar}, \menv), \mstore, \mframe, \mkont}$
    &
    $\tpl{\mval,\mstore,\mframe',\mkont}$ if $(\mframe', \mval) \in \lookup(\mstore,\mframe,\menv,\mvar,\mlab)$
    \\
% % application
%     $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mframe, \mkont}$
%     &
%     $\tpl{(\mexpri0, \menv), \mstore, \mframe, \kar{\mexpri1,\menv,\mkont}}$
%     \\
% % argument evaluation
%     $\tpl{\mval, \mstore, \mframe, \kar{\mexpr,\menv,\mkont}}$
%     &
%     $\tpl{(\mexpr, \menv), \mstore, \mframe, \kfn{\mval, \mkont}}$
%     \\
% function call
    $\tpl{\mval,\mstore,\mframe,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \mstore', \mframe', \kcons{\kpush{\mframe}}{\mkont}}$
    \\ where & $(\mstore',\mframe') = \bind(\mstore,\mframe,\maddr,\mvar,\mval)$
    \\
    $\tpl{\mval,\mstore,\mframe,\kcons{\kpush{\mframe'}}{\mkont}}$
    &
    $\tpl{\mval,\mstore,\mframe',\mkont}$
  \end{tabular}
  \caption{The CES$\xi$K machine}
  \label{fig:frame-semantics}
\end{figure}

The meta-functions the semantics uses to extend and consult the heap
and stack use implicit information from the pre-analysis we described
above:

\begin{align*}
  \bind(\mstore,\mframe,\maddr,\mvar,\mval) &=
   \left\{\begin{array}{l}
            (\mstore, \snglm{\maddr}{\mval}) \text{ if } \mvar \text{ never escapes} \\
            (\joinone{\mstore}{\maddr}{\mval}, \snglm{\maddr}{\mval}) \text{ otherwise} \\
          \end{array}\right. \\
  \lookup(\mstore,\mframe,\menv,\mvar,\mlab) &=
    \left\{\begin{array}{l}
          \setbuild{(\extm{\mframe}{\menv(\mvar)}{\set{\mval}}, \mval)}
                   {\mval \in \mframe(\menv(\mvar))}
                   \\ \qquad \text{if } \mlab \text{ non-escaping} \\
          \setbuild{(\mframe,\mval)}{\mval \in \mstore(\menv(\mvar))} \text{ otherwise}
           \end{array}\right.
\end{align*}

@@

%%
CFA2 also includes what they called ``transitive summaries'' to deal with tail calls.
%
We insert an $\rt$ frame at every function call.
%
This view contends with tail calls, but we can identify tail calls easily --- any call with an $\rt$ as its continuation is a tail call.
%
Tail calls are important in language implementations for space complexity reasons \citep{ianjohnson:clinger:tail-calls:1998}, but in an analysis, these concerns are less important.
%
The repeated popping of $\rt$ inserted by what otherwise were tail-calls is synonymous with CFA2's transitive summaries.

@@

Figure \ref{fig:shift-reset-table0} has what one might naturally write using just this information.
%
Unfortunately, since $\rt$ frames contain a store, and continuations can now appear in the store, this introduces a circularity that could cause the analysis to never terminate.
%
For example, a loop that continually captures continuations would introduce unboundedly many continuation values due to the ever-growing store.
%
This means the store is not finite height, and there may be no fixed point.

\paragraph{Possible fixes to non-termination:} The simplest fix to the circularity would be to strip stores out of the context of the $\rt$ frame in a captured continuation, and additionally consider contexts with a stripped store an abstraction of the same context with any store.
%
This is a fairly brutal approximation of a captured continuation, so an alternative is to make this tunable with our precision-tuning friend, $\alloc$.
%
Then, instead of entirely removing the store component of the $\rt$ context, we can replace it with an address of possible stores that it would approximate.

@@

The meta-functions mentioned in the fixed semantics all deal with the addition of $\maddr$ to contexts.
%
If the context in an $\rt$ frame is approximate, we must return to all the continuations known for all the contexts it approximates:
\begin{align*}
  \returns(\mktab, ((\mexpr,\menv), \mstore)) &= \mktab((\mexpr,\menv), \mstore) \\
  \returns(\mktab, ((\mexpr,\menv), \maddr)) &=
    \bigcup\setbuild{\mktab((\mexpr,\menv),\mstore)}{\mstore \in \mktab(\maddr)}
\end{align*}

At return boundaries, the memo table must add the result to all the represented contexts:=
\begin{align*}
  \memoize(\mktab, ((\mexpr,\menv), \mstore'), \mathit{v\sigma}) &=
  [((\mexpr,\menv), \mstore') \mapsto \set{{\mathit{v\sigma}}}] \\
  \memoize(\mktab, ((\mexpr,\menv), \maddr), \mathit{v\sigma}) &=
  \bigsqcup\limits_{\mstore' \in \mktab(\maddr)}\snglm{((\mexpr,\menv), \mstore')}{\mathit{v\sigma}}
\end{align*}
At capture time, we strip the store in the $\rt$ frame if there is one, and replace it with an address:

>>>>

%%
There is a systematic process for transforming off-the-shelf programming language semantics into a form amenable to \emph{regular} analysis that has been widely applied with great success to production programming languages~\citet{dvanhorn:VanHorn2010Abstracting} (a technique called abstracting abstract machines, or AAM).
%
A contribution of this paper is a systematic process to construct \emph{pushdown} analyses of programming languages, due to the precision (and often performance) benefits.
%%

@@

The remaining sections of the paper are
\begin{itemize}
\item{\autoref{sec:pdcfa}: we derive a cousin of PDCFA}
\item{\autoref{sec:gc}: we show how abstract garbage collection ``falls out'' of this approach.}
\item{\autoref{sec:cfa2}: we make additions to the PDCFA semantics to get a direct-style CFA2 without first-class control}
\item{\autoref{sec:sr}: we give a novel analysis of delimited and composable first-class control.}
\end{itemize}

@@

%%
The tail of a continuation in the context of a function will contain the stack-less context in which the function was called, in order to link up with the proper call-site(s).
%
The context includes the function (or unique label of the function), the environment, and the store.
%
Notice that this choice can be viewed as a special allocation strategy for continuations in the AAM viewpoint, but we separate the table of continuations from the store since continuations themselves contain stores in $\rt$ continuations --- this leads to a recursive data-structure that we are trying to avoid.
%
The store is an important ingredient to maintaining enough precision to keep a pushdown abstraction.
%
The semantic spaces for the machine are modified thusly:

@@

The $\mathit{unroll}$ function interprets what are all the valid continuations that $\mktab$ encodes for a given continuation that contains $\krt{\mctx}$, defined as the greatest fixed point of the following rules:
\begin{mathpar}
  \inferrule{ }{\kmt \in \unroll{\mktab}{\kmt}} \quad
  \inferrule{\mkont \in \unroll{\mktab}{\mkont'}}{\mkframe:\mkont \in \unroll{\mktab}{\mkframe:\mkont'}} \quad
  \inferrule{\mkont^\circ \in \mktab(\mctx) \\
             \mkont \in \unroll{\mktab}{\mkont^\circ}}
            {\mkont \in \unroll{\mktab}{\krt{\mctx}}}
\end{mathpar}

The astute reader may find this definition is dubious, since there is a possibility for unguarded corecursion here.
%
Interpreting a function with ill-founded recursion can lead to a table such that $\krt{\mctx} \in \mktab(\mctx)$.
%
The abstractions that we impose on our data might also cause this situation, even though the concrete execution might always terminate.
%
Our interpretation of $\stepto$ is in a fixed-point computation, so circularities are not a problem for termination, but we augment the definition here to guard against circularities in our proofs:

\begin{mathpar}
  \inferrule{ }{\kmt \in \unroll{\mktab}{\kmt, G}} \quad
  \inferrule{\mkont \in \unroll{\mktab}{\mkont', \bot}}{\mkframe:\mkont \in \unroll{\mktab}{\mkframe:\mkont', G}} \\
  \inferrule{G \neq \mctx \\
             \mkont^\circ \in \mktab(\mctx) \\
             \mkont \in \unroll{\mktab}{\mkont^\circ, \mctx}}
            {\mkont \in \unroll{\mktab}{\krt{\mctx}, G}}  
\end{mathpar}
Consider later uses of $\mathit{unroll}$ to implicitly give $\bot$ for $G$.

@@

\section{Deriving GC from the introspective pushdown model}\label{sec:gc}

@@

This semantics of course enjoys the soundness properties that GC does not introduce dangling pointers, and (in the concrete) any two reduction sequences of the same length lead to states equal up to GC.
%
Unconstrained allocation functions

@@

CFA2 is the first published analysis of a higher-order programming language that could properly match calls and returns.
%
We will show that it fits well into the same presentation we gave for PDCFA.
%
Vardoulakis and Shivers had a clear goal of harnessing the extra information a pushdown model provides to produce a high-precision analysis that works well in practice.
%
This resulted in more than just the call/return matching of the previous section, which is why we are showing the two separately.
%
There are two orthogonal features of the semantics, embodied by our semantics' two metafunctions ($\bind$ and $\lookup$, respectively in \autoref{fig:frame-semantics}):
\begin{enumerate}
\item{stack allocation for some bindings in an additional $\mframe \in \Store$}
\item{strong updates on stack frames for resolved non-determinism}
\end{enumerate}
The first of these is an addition to the stack-less context.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar[\mlab]{\mvar}, \menv), \mstore, \mframe, \mkont}$
    &
    $\tpl{\mval,\mstore,\mframe',\mkont}$ if $(\mframe', \mval) \in \lookup(\mstore,\mframe,\menv,\mvar,\mlab)$
    \\
% % application
%     $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mframe, \mkont}$
%     &
%     $\tpl{(\mexpri0, \menv), \mstore, \mframe, \kar{\mexpri1,\menv,\mkont}}$
%     \\
% % argument evaluation
%     $\tpl{\mval, \mstore, \mframe, \kar{\mexpr,\menv,\mkont}}$
%     &
%     $\tpl{(\mexpr, \menv), \mstore, \mframe, \kfn{\mval, \mkont}}$
%     \\
% function call
    $\tpl{\mval,\mstore,\mframe,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \mstore', \mframe', \kcons{\kpush{\mframe}}{\mkont}}$
    \\ where & $(\mstore',\mframe') = \bind(\mstore,\maddr,\mvar,\mval)$
    \\
    $\tpl{\mval,\mstore,\mframe,\kcons{\kpush{\mframe'}}{\mkont}}$
    &
    $\tpl{\mval,\mstore,\mframe',\mkont}$
  \end{tabular}
  \caption{The CES$\xi$K machine}
  \label{fig:frame-semantics}
\end{figure}

%
There is a conservative pre-analysis that checks locally whether a binding will never escape, and classifies references (labeled with a distinguishing $\mlab$ from an arbitrary space of labels) as able to use the stack frame or not.
%
Their criteria for a binding never escaping is that it is never referenced in a function that is not its binder.
%
This can be extended in a language with more linguistic features; see Kranz's thesis about register-allocatable bindings in the Orbit Scheme compiler \citep{ianjohnson:kranz:thesis:1988}.
%
A stackable reference is one that appears in the body of the binding function, by which we mean not within the body of a nested function.
%
They use the information that a binding never escapes to not bother allocating it in the heap.
%
This has the advantage of not changing the heap, and thus leads to less propagation.
%
The addition of these stack frames makes the analysis exponential in theory, though in practice they have been observed to decrease running time in most cases.
%%

@@

%%
We show only the significantly modified rules of the semantics in figure \ref{fig:frame-semantics}.
%
The other rules simply carry along the extra $\mframe$ component untouched.
%
The meta-functions the semantics uses to extend and consult the heap and stack use implicit information from the pre-analysis we described above:

\begin{align*}
  \bind(\mstore,\maddr,\mvar,\mval) &=
   \left\{\begin{array}{l}
            (\mstore, \snglm{\maddr}{\mval}) \text{ if } \mvar \text{ allocations never escape} \\
            (\joinone{\mstore}{\maddr}{\mval}, \snglm{\maddr}{\mval}) \text{ otherwise} \\
          \end{array}\right. \\
  \lookup(\mstore,\mframe,\menv,\mvar,\mlab) &=
    \left\{\begin{array}{l}
          \setbuild{(\extm{\mframe}{\menv(\mvar)}{\set{\mval}}, \mval)}
                   {\mval \in \mframe(\menv(\mvar))}
                   \\ \qquad \text{if } \mlab \text{ non-escaping} \\
          \setbuild{(\mframe,\mval)}{\mval \in \mstore(\menv(\mvar))} \text{ otherwise}
           \end{array}\right.
\end{align*}

@@

A tabular semantics for the CES$\xi$K machine follows the previous section almost entirely, except that $\mktab$ now stores $(\mkont, \mframe)$ pairs, and returns reinstate the saved $\mframe$:
\begin{align*}
  \tpl{\mval,\mstore,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont},\mktab,\mmemo} &\stepto
  \tpl{\mpoint,
    \mstore',
    \mframe',
    \krt{\mpoint, \mstore'},
    \mktab',
    \mmemo} \\
  \text{where }
  & \mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr}) \\
  & (\mstore',\mframe') = \bind(\mstore,\maddr,\mvar,\mval) \\
  & \mktab' = \joinone{\mktab}{(\mpoint, \mstore')}{(\mkont,\mframe)}
\\[2pt]
 \tpl{\mval, \mstore, \mframe, \krt{\mpoint,\mstore'}, \mktab, \mmemo} &\stepto
 \tpl{\mval, \mstore, \mframe', \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{(\mval,\mstore)}}
  \\ & \quad\text{if } (\mkont,\mframe') \in \mktab(\mpoint, \mstore')
\end{align*}

%%
CFA2 also includes what they called ``transitive summaries'' to deal with tail calls.
%
We create an $\rt$ continuation at every function call.
%
This view contends with tail calls, but we can identify tail calls easily --- any call with an $\rt$ as its continuation is a tail call.
%
Tail calls are important in language implementations for space complexity reasons \citep{ianjohnson:clinger:tail-calls:1998}, but in an analysis, these concerns are less important.
%
The repeated popping of $\rt$ inserted by what otherwise were tail-calls is synonymous with CFA2's transitive summaries.

@@

Figure \ref{fig:shift-reset-table0} has what one might naturally write using just this information.
%
Unfortunately, since $\rt$ continuations contain a store, and continuations can now appear in the store, this introduces a circularity that could cause the analysis to never terminate.
%
For example, a loop that continually captures continuations would introduce unboundedly many continuation values due to the ever-growing store.
%
This means the store is not finite height, and there may be no fixed point.

\paragraph{Possible fixes to non-termination:} The simplest fix to the circularity would be to strip stores out of the context of the $\rt$ in a captured continuation, and additionally consider contexts with a stripped store an abstraction of the same context with any store.
%
This is a fairly brutal approximation of a captured continuation, so an alternative is to make this tunable with our precision-tuning friend, $\alloc$.
%
Then, instead of entirely removing the store component of the $\rt$ context, we can replace it with an address of possible stores that it would approximate.

@@

The meta-functions mentioned in the fixed semantics all deal with the addition of $\maddr$ to contexts.
%
If the context in an $\rt$ continuation is approximate, we must return to all the continuations known for all the contexts it approximates:
\begin{align*}
  \returns(\mktab, ((\mexpr,\menv), \mstore)) &= \mktab((\mexpr,\menv), \mstore) \\
  \returns(\mktab, ((\mexpr,\menv), \maddr)) &=
    \bigcup\setbuild{\mktab((\mexpr,\menv),\mstore)}{\mstore \in \mktab(\maddr)}
\end{align*}

At return boundaries, the memo table must add the result to all the represented contexts:=
\begin{align*}
  \memoize(\mktab, ((\mexpr,\menv), \mstore'), \mathit{v\sigma}) &=
  [((\mexpr,\menv), \mstore') \mapsto \set{{\mathit{v\sigma}}}] \\
  \memoize(\mktab, ((\mexpr,\menv), \maddr), \mathit{v\sigma}) &=
  \bigsqcup\limits_{\mstore' \in \mktab(\maddr)}\snglm{((\mexpr,\menv), \mstore')}{\mathit{v\sigma}}
\end{align*}
At capture time, we strip the store in the $\rt$ continuation if there is one, and replace it with an address:
